---
title: "VisualizeSPECInt"
author: "Winston"
date: "September 26, 2014"
output: 
html_document:
    keep_md: true
---

###SPEC cpu2006 Trends for PCs

###Synopisis

This is a quick look at trends in CPU performance based on the set of published SPECCPU2006 benchmark. These are published at the [SPEC.org](http://spec.org/cpu2006/results/) website where you can read more about the benchmarkand data collection.

CPUInt, studied here, is an important characteristic of computer performance. It measures the speed of integer (as opposed to floating point) operations.


The question addressed here is "what factors have contributed most to increases in SPU performance?"


###Data Processing

The program checks for the existence of SPECCPU2006.csv on the local drive. If it does not find the file, it downloads it and then reads it into memory. 
The file contains about 30,000 observations.  
The data as downloaded are already very clean. The only real modifications are:  
- date columns translated to _R_ data format.  
- eliminate some zero Result values  

```{r "get_data", echo=TRUE, cache=TRUE}


## data are located at the SPEC website
url <- "http://spec.org/cgi-bin/osgresults?conf=cpu2006;op=dump;format=csvdump"
destfile <- "SPECCPU2006.csv"

## check if the data are stored locally, and if not download
if(!file.exists(destfile)) download.file(url, destfile)

##read the data as a .csv file
SPECdata <- read.csv(destfile)
## the structure of the raw data
str(SPECdata)

```

```{r "data_clean_dates", cache=TRUE}

##convert data to dates in R format
        ##The incoming data are in a mmm-YYYY format. 
        ##I had to append a day to the date data (which I chose as the 15th (arbitrarily))
        ## All date data are scrubbed though not all are necessarily used.

        ## HW.Avail data
        SPECdata$HW.Avail <- as.character(SPECdata$HW.Avail)
        SPECdata$HW.Avail <- paste0("15-", SPECdata$HW.Avail)
        SPECdata$HW.Avail<- as.Date(SPECdata$HW.Avail, format="%d-%b-%Y")

        ## SW.Avail date
        SPECdata$SW.Avail <- as.character(SPECdata$SW.Avail)
        SPECdata$SW.Avail <- paste0("15-", SPECdata$SW.Avail)
        SPECdata$SW.Avail<- as.Date(SPECdata$SW.Avail, format="%d-%b-%Y")

        ## Test Date
        SPECdata$Test.Date <- as.character(SPECdata$Test.Date)
        SPECdata$Test.Date <- paste0("15-", SPECdata$Test.Date)
        SPECdata$Test.Date<- as.Date(SPECdata$Test.Date, format="%d-%b-%Y")

        ## Publication date
        SPECdata$Published <- as.character(SPECdata$Published)
        SPECdata$Published <- paste0("15-", SPECdata$Published)
        SPECdata$Published<- as.Date(SPECdata$Published, format="%d-%b-%Y")

```

```{r "data_clean2_Values"}

##clean up results 
## Some spurious zero values need to be eliminated.
SPECdata<-SPECdata[SPECdata$Result>3,]


```


Here is a look at some of the data.   To simplify the plot this data has been filtered for systems with two chips and the data factored for cores per chip.  
Note the log scale. Roughly speaking computing performance has increased over a factor of ten since 2006, when the benchmark was first introduced.  
Note that the data split into two bands.  We'll explore this in the Results section.

```{r "expository_plot", echo=TRUE, cache=FALSE, fig.align='center'}

#Plot Data as function of date by various facets

library(ggplot2)

SPECdataX<-SPECdata[SPECdata$X..Chips.==2,]

P1 <- ggplot(SPECdataX, aes(x=HW.Avail, y=Result))+geom_point(shape=1)
P1 <- P1 + scale_y_log10()
P1<- P1+ geom_point(aes(colour = factor(X..Cores.Per.Chip.)))
#P1 <- P1 + facet_grid(X..Cores ~.)

print(P1)



```

###Results

```{r "Intel_Core_and_Xeon", echo=TRUE,fig.width=8,fig.align='center' }

##Define ComputerType_x function to reduce data based on substring
        
        ##name_string is the string used to catergorize results (e.g. "Core")
        ##search string is what is used to select elements for the category
        ##Output is the data aggregation file. Note it should be set to NULL prior to running this for the first time.
        ##input is the raw data file
        ##The Outputfile add a column to the input data with the output 

        ComputerType_x <- function(name_string, search_string, output=SPECdata_x, input=SPECdata) { 
                temp <- input[grepl(search_string, input$Processor., ignore.case=TRUE), ]
                names <- rep(name_string, dim(temp)[1])
                temp<- cbind (names, temp)
                rbind(temp, output)
                }

SPECdataReduce<-SPECdata[SPECdata$X..Chips.==1,]

SPECdata_x<-NULL

        SPECdata_x<-ComputerType_x("Intel Core", "core", input=SPECdataReduce)
        SPECdata_x<-ComputerType_x("Intel Xeon", "xeon")

P1 <- ggplot(SPECdata_x, aes(x=HW.Avail, y=Result))+geom_point(shape=1)
P1 <- P1 + scale_y_log10()
P1<- P1+ geom_point(aes(colour = factor(X..Cores.Per.Chip.)))
P1<- P1 + facet_grid(. ~ names)
#P1 <- P1 + facet_grid(X..Cores ~.)

print(P1)

str(SPECdata_x)